---
title: "Severe U.S. weather events affecting health and economy"
author: "Jessica"
date: "11 juli 2018"
output: html_document
---


## Synopsis

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe weather events can result in fatalities, injuries, and property damage. Preventing such outcomes to the extent possible is a key concern. It is therefore essential to know across the United States:

* which types of events are most harmful with respect to population health
* which types of events have greatest economic consequesces

These answers are obtained by analysing the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. We can conclude that in preparation for severe weather events, prirority must be given to resources for the following types of weather events:

* Tornado/hurricane/typhoon
* Floods
* Excessive heat



## Data Processing

The registerations in the NOAA started in the year 1950 and ended in November 2011. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage. This data is provided in csv and loaded in R.

```{r load_data, echo = TRUE, cache = TRUE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
storm_data <- read_csv("repdata%2Fdata%2FStormData.csv.bz2")
```

The impact of weather events on population health will be measured by looking at both the number of fatalities and injuries. For determening the property damage, only the "correct" registrations are included. These are the values for PROPDMG where the PROPDMGEXP has values of "K", "M", or "B".

So after loading the raw data in R, the following data preparations are done:
- selecting only the columns required for this analysis
- transforming the property damage into usable numerics/valuta

```{r data_prep, echo=TRUE, message=FALSE}
analysis_data <- storm_data %>%
  select(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP) %>%
  mutate(value_property_damage = ifelse(PROPDMGEXP == "K", PROPDMG * 1000, 
                                        ifelse(PROPDMGEXP %in% c("M","m"), PROPDMG * 1000000,
                                               ifelse(PROPDMGEXP == "B", PROPDMG * 1000000000, 0))))
  
```



## Results

The following graph shows which types of events are most harmful with respect to population health. Tornado's cause bay far the injuries and fatalities.

```{r plot_social_impact, echo=TRUE, message=FALSE}

analysis_data %>%
  group_by(EVTYPE)%>%
  summarise(total_fatalities = sum(FATALITIES, na.rm=TRUE),
            total_injuries = sum(INJURIES, na.rm = TRUE),
            total = total_fatalities+total_injuries) %>%
  ungroup() %>%
  arrange(desc(total)) %>%
  head(n=7) %>%
  select(-total) %>%
  gather(key = "injuries_fatalities", value = "value", -EVTYPE) %>%
ggplot(aes(x = reorder(EVTYPE, value), y = value, fill = injuries_fatalities))+
  geom_col()+
  ggtitle("Total injuries and fatalities in US", subtitle = "For top 7 most harmful weather events")+
  labs(x = "weather events", y = "number of injuries and fatalities")+
  coord_flip()
  
```


As shown in the following graph, flood and wind related weather events (tornado, hurricane, storm) cause most damage to properties in the US. 

```{r plot_economic_impact, echo=TRUE, message=FALSE}
analysis_data %>%
  group_by(EVTYPE) %>%
  summarise(total_damage = sum(value_property_damage, na.rm = TRUE))%>%
  ungroup() %>%
  arrange(desc(total_damage)) %>%
  head(n=7) %>%
ggplot(aes(x = reorder(EVTYPE, total_damage), y = total_damage)) +
  geom_col() +
  ggtitle("Total property damage in US", subtitle = "For top 7 most damaging weather events") +
  labs(x ="weather events", y = "estimated property damage in USD ($)") +
  coord_flip()
  
```

Looking at these results, we can conclude that resources should be aimed at the following severe weather events:

* Tornado/hurricane/typhoon
* Floods
* Excessive heat
