---
title: "Storm Data"
author: "Jessica"
date: "6 juli 2018"
output: html_document
---


Your data analysis must address the following questions:
- Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
- Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.


## Synopsis

Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.


## Data Processing

There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data.

```{r load_data, echo = TRUE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
storm_data <- read_csv("repdata%2Fdata%2FStormData.csv.bz2")
```


## Results

There should be a section titled Results in which your results are presented. You may have other sections in your analysis, but Data Processing and Results are required. The analysis document must have at least one figure containing a plot. Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

```{r plot_injuries, echo=TRUE}

storm_data %>%
  group_by(EVTYPE)%>%
  summarise(total_fatalities = sum(FATALITIES, na.rm=TRUE),
            total_injuries = sum(INJURIES, na.rm = TRUE),
            total = total_fatalities+total_injuries) %>%
  arrange(desc(total))%>%
  head(n=8)%>%
  select(-total)%>%
  gather(key = "injuries_fatalities", value = "value", -EVTYPE)%>%
ggplot(aes(x=EVTYPE, y=value, fill = injuries_fatalities))+
  geom_col()+
  ggtitle("Total injuries and fatalities per event type")+
  coord_flip()
  
```

